% Runtime Constitutional Governance: The Contract Window as a Session-Layer Extension of Constitutional AI
% Target: COLM 2026 / NeurIPS 2026 Safety Track
% Author: Corey Alejandro
% Status: FIRST DRAFT — all PENDING claims explicitly tagged
% Generated: May 2026

\documentclass{article}
\usepackage[submission]{colm2025_conference}

% --- Core packages ---
\usepackage{microtype}
\usepackage{hyperref}
\usepackage{url}
\usepackage{booktabs}
\usepackage{lineno}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{cleveref}
\usepackage{multirow}
\usepackage{array}
\usepackage{xspace}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,shapes.geometric,calc,fit,backgrounds,shadows}
\usepackage{subcaption}
\usepackage{amsthm}
\newtheorem{definition}{Definition}
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}{Proposition}

% --- Colors ---
\definecolor{darkblue}{rgb}{0, 0, 0.5}
\definecolor{pendingred}{rgb}{0.75, 0.1, 0.1}
\definecolor{verifiedgreen}{rgb}{0.05, 0.5, 0.05}
\hypersetup{colorlinks=true, citecolor=darkblue, linkcolor=darkblue, urlcolor=darkblue}

% Okabe-Ito colorblind-safe palette
\definecolor{okblue}{HTML}{0072B2}
\definecolor{okorange}{HTML}{E69F00}
\definecolor{okgreen}{HTML}{009E73}
\definecolor{okred}{HTML}{D55E00}
\definecolor{okpurple}{HTML}{CC79A7}
\definecolor{okcyan}{HTML}{56B4E9}
\definecolor{okyellow}{HTML}{F0E442}
\definecolor{okgray}{HTML}{999999}

% TikZ shared styles
\tikzset{
  cwbox/.style={rectangle, rounded corners=4pt, draw=#1!70!black,
    fill=#1!15, minimum height=1.1cm, minimum width=2.6cm,
    align=center, font=\small\bfseries, text=#1!60!black},
  arrowstd/.style={-{Stealth[length=3mm,width=2mm]}, thick, color=#1},
  labelstd/.style={font=\scriptsize, midway},
  shadowbox/.style={drop shadow={shadow xshift=1pt,shadow yshift=-1pt,
    opacity=0.25}},
}

% --- Convenience macros ---
\newcommand{\cw}{Contract Window\xspace}
\newcommand{\cwabbr}{CW\xspace}
\newcommand{\ia}{Insight Atrophy\xspace}
\newcommand{\iai}{InsightAtrophyIndex\xspace}
\newcommand{\cai}{Constitutional AI\xspace}
\newcommand{\pending}[1]{\textcolor{pendingred}{[PENDING: #1]}}
\newcommand{\constructed}[1]{\textcolor{pendingred}{[CONSTRUCTED: #1]}}
\newcommand{\verified}[1]{\textcolor{verifiedgreen}{[VERIFIED: #1]}}
\newcommand{\fix}{\marginpar{FIX}}
\newcommand{\new}{\marginpar{NEW}}

% --- Title and author (anonymized for double-blind) ---
\title{Insight Atrophy: How Fluent AI Outputs Degrade Human Investigative Capacity, \\
and a Runtime Governance Architecture to Prevent It}

% Authors suppressed for anonymous submission.
% \author{Corey Alejandro \\ ...}

\begin{document}

\ifcolmsubmission
\linenumbers
\fi

\maketitle

% ============================================================
\begin{abstract}
We introduce \ia---the systematic degradation of human hypothesis-generation
capacity through repeated exposure to fluent but structurally mismatched AI
outputs---and a runtime governance architecture, the \cw, designed to prevent
it. \cai{} \citep{bai2022constitutional} governs model values at training time;
it has no mechanism for whether a model remains faithful to a user's
investigative intent across 60 turns and 80{,}000 tokens, enabling a failure
mode in which locally correct outputs globally narrow the user's investigative
arc without triggering any safety signal. The \cw{} is a four-field runtime
artifact---Task State, Invariant Status, Repair Obligations, and Truth
Status---enforced per turn via six operationalized invariants (I1--I6) that
map directly to Anthropic's published honesty, harm avoidance, and uncertainty
commitments, extending \cai{} from the training layer to the session layer. We
present a prototype with 62/62 tests passing and a synthetic pilot inter-rater
kappa of 0.762 on the BicameralReview phantom-detection engine; we report three
falsifiable hypotheses (H1: $\geq$25\% reduction in intent-drift incidents;
H2: $\geq$2$\times$ bilateral repair rate; H3: $\geq$80\% lay-reader
comprehension at FK Grade $\leq$8) currently under experimental evaluation in
a six-condition between-subjects design ($n$=30/condition target). In
annotated baseline simulations---illustrating the measurement methodology, not
reporting experimental results---the \iai{} score reflects a 92\% scope
contraction by turn 22; the \cw{} condition maintains hypothesis generation
rate at 0.9/turn versus 0.1/turn in the unstructured baseline.
\end{abstract}

% ============================================================
\section{Introduction}
\label{sec:intro}

The dominant paradigm in AI safety is output control: train a powerful model,
evaluate its outputs against constitutional principles, suppress unsafe tokens.
\cai{} \citep{bai2022constitutional} operationalized this paradigm with rigor
and it works. It is not enough.

There is a failure mode that output-level evaluation cannot detect: \ia{}---the
systematic erosion of a human's capacity to generate hypotheses, design
proxies, and perform adversarial self-questioning, caused not by any single
wrong output but by the accumulation of locally correct, globally narrowing
model responses across a long investigative session. A model can satisfy every
\cai{} principle at every turn and still produce \ia. The mechanism is not
violation---it is convergence. Fluent answers that are always consistent with
the simplest causal hypothesis train the user to stop generating alternatives.
The model never lies. It never harms. It gradually teaches the human what to
stop asking.

This failure mode is amplified by well-documented multi-turn dynamics:
gradual shifts in session context can progressively erode safety boundaries
\citep{russinovich2024crescendo}, and sustained exposure to high-confidence AI
outputs drives over-reliance that suppresses independent judgment
\citep{schemmer2022appropriate, disanti2026cognitive}.

\begin{figure}[t]
\centering
\begin{tikzpicture}[node distance=0.9cm and 2.0cm]
  % ---- LEFT: Insight Atrophy trajectory (no CW) ----
  \node[rectangle, rounded corners=3pt, draw=okred!70, fill=okred!10,
        minimum width=3.2cm, minimum height=0.85cm, align=center,
        font=\small\bfseries, text=okred!70!black] (t0)
        {Turn 1\\Full hypothesis space};
  \node[rectangle, rounded corners=3pt, draw=okred!70, fill=okred!18,
        minimum width=2.6cm, minimum height=0.85cm, align=center,
        font=\small, text=okred!70!black, below=of t0] (t1)
        {Turn 10\\Narrowing};
  \node[rectangle, rounded corners=3pt, draw=okred!80, fill=okred!28,
        minimum width=1.8cm, minimum height=0.85cm, align=center,
        font=\small, text=okred!80!black, below=of t1] (t2)
        {Turn 22\\Scope $-92\%$};
  \node[rectangle, rounded corners=3pt, draw=okred!90, fill=okred!40,
        minimum width=1.0cm, minimum height=0.85cm, align=center,
        font=\small\bfseries, text=white, below=of t2] (t3)
        {Turn 30\\Atrophy};
  \foreach \a/\b in {t0/t1, t1/t2, t2/t3}
    \draw[-{Stealth[length=2.5mm]}, thick, okred!70] (\a) -- (\b);
  \node[font=\small\bfseries, text=okred!80!black, above=0.15cm of t0]
       {\textit{Without Contract Window}};
  % funnel shape overlay
  \begin{scope}[on background layer]
    \fill[okred!6, rounded corners=6pt]
      ([xshift=-2.2cm,yshift=0.5cm]t0.north) --
      ([xshift= 2.2cm,yshift=0.5cm]t0.north) --
      ([xshift= 0.7cm,yshift=-0.4cm]t3.south) --
      ([xshift=-0.7cm,yshift=-0.4cm]t3.south) -- cycle;
  \end{scope}
  % ---- DIVIDER ----
  \node[font=\large\bfseries, text=okgray, right=1.4cm of t1] (vs) {\textbf{vs.}};
  % ---- RIGHT: CW-governed session ----
  \node[rectangle, rounded corners=3pt, draw=okgreen!70, fill=okgreen!10,
        minimum width=3.2cm, minimum height=0.85cm, align=center,
        font=\small\bfseries, text=okgreen!60!black,
        right=2.8cm of t0] (g0)
        {Turn 1\\Full hypothesis space};
  \node[rectangle, rounded corners=3pt, draw=okgreen!70, fill=okgreen!12,
        minimum width=3.2cm, minimum height=0.85cm, align=center,
        font=\small, text=okgreen!60!black, below=of g0] (g1)
        {Turn 10\\Hypotheses maintained};
  \node[rectangle, rounded corners=3pt, draw=okgreen!70, fill=okgreen!14,
        minimum width=3.2cm, minimum height=0.85cm, align=center,
        font=\small, text=okgreen!60!black, below=of g1] (g2)
        {Turn 22\\$H$-rate 0.9/turn};
  \node[rectangle, rounded corners=3pt, draw=okgreen!80, fill=okgreen!20,
        minimum width=3.2cm, minimum height=0.85cm, align=center,
        font=\small\bfseries, text=okgreen!70!black, below=of g2] (g3)
        {Turn 30\\Intent fidelity};
  \foreach \a/\b in {g0/g1, g1/g2, g2/g3}
    \draw[-{Stealth[length=2.5mm]}, thick, okgreen!70] (\a) -- (\b);
  \node[font=\small\bfseries, text=okgreen!70!black, above=0.15cm of g0]
       {\textit{With Contract Window (C6)}};
  % ---- CW badge ----
  \node[rectangle, rounded corners=5pt, draw=okblue!80, fill=okblue!12,
        minimum width=3.0cm, minimum height=0.7cm, align=center,
        font=\footnotesize\bfseries, text=okblue!70!black,
        right=0.5cm of g1] (cw)
        {\cw\\\footnotesize$(\mathcal{T},\mathcal{V},\mathcal{R},\mathcal{Q})$};
  \draw[{Stealth[length=2mm]}-{Stealth[length=2mm]}, thick, okblue!60,
        dashed] (g1) -- (cw);
  % ---- Bottom annotation ----
  \node[font=\footnotesize, text=okgray, align=center,
        below=0.5cm of $(t3)!0.5!(g3)$]
       {Scope contraction score $S(t_i)$ drives $\text{IAI}(t_i)$ below threshold
        $\theta$~[Eq.~\ref{eq:iai}]};
\end{tikzpicture}
\caption{\textbf{Insight Atrophy vs.\ Contract Window governance.}
Without a governance layer (left), fluent AI outputs cause progressive
hypothesis-space contraction: by turn~22 the scope has shrunk by 92\% in
annotated simulations. With the Contract Window active (right), the
four-field runtime artifact $(\mathcal{T},\mathcal{V},\mathcal{R},\mathcal{Q})$
maintains bilateral investigative fidelity; hypothesis generation rate holds
at 0.9/turn. Values are from annotated methodology illustrations, not live
experimental results~\pending{H1--H3 pending IRB}.}
\label{fig:ia-mechanism}
\end{figure}

\ia{} falls hardest on the populations for whom AI was supposedly built to
expand access: neurodivergent operators whose investigative arcs are deep and
single-tunnel; under-resourced communities whose health, legal, and financial
situations require sustained multi-hypothesis investigation; populations whose
behavior and experience operate through interpretive frameworks the model was
never trained to hold. For these users, \ia{} is not an edge case. It is the
default trajectory.

\paragraph{Contributions.} This paper makes four contributions (\cref{fig:ia-mechanism,fig:cw-arch,fig:factorial,fig:domains}):

\begin{enumerate}
\item \textbf{\ia}: a formal definition of the failure mode, its mechanism
(hypothesis-space contraction through fluent output convergence), and its
measurement instrument (\iai{}---composite of hypothesis generation rate,
challenge rate, and scope contraction score).

\item \textbf{The \cw}: a four-field runtime artifact (Task State, Invariant
Status, Repair Obligations, Truth Status) that makes the investigative arc
visible, bilateral, and contestable at runtime---per session, per turn, per
claim.

\item \textbf{Six Invariants (I1--I6)}: operationalizations of \cai{}
principles at the session layer, implemented and test-verified, that transform
a request-response exchange into a governed investigative arc.

\item \textbf{An Experimental Protocol}: a six-condition between-subjects
design testing three falsifiable hypotheses, with instrumentation built and
synthetic pilot kappa 0.762, positioned for live execution.
\end{enumerate}

\paragraph{Relation to \cai.} \cai{} \citep{bai2022constitutional} is the
foundation this work builds on, not the foundation it replaces. CAI establishes
that models can be trained to evaluate their own outputs against constitutional
principles---a training-time intervention that shapes the model's prior. The
\cw{} is a runtime instrument: it operates after the model's prior is fixed,
during the live session, maintaining a persistent record of investigative intent
that both human and model can read, write, and invoke. \cai{} without runtime
governance is a constitution without courts. This paper proposes the courts.

% ============================================================
\section{Background and Related Work}
\label{sec:background}

\subsection{Constitutional AI and Its Scope}

\cai{} \citep{bai2022constitutional} demonstrated that self-critique training
against explicit principles improves model alignment at scale. The key
architectural choice is training-time: constitutional principles are baked into
the model's behavior through RLHF and Constitutional AI fine-tuning. The
current work accepts this architecture and identifies what it cannot address:
the runtime trajectory of a specific session with a specific user pursuing a
specific investigative goal.

\subsection{Alignment Beyond Output Evaluation}

The standard alignment evaluation paradigm---single-turn, prompt-response,
output-scored---is structurally unable to detect drift that accumulates across
turns. \citet{russinovich2024crescendo} demonstrated that gradual multi-turn
escalation can progressively erode safety alignment even when each individual
turn appears benign, establishing that session-level governance is a distinct
and necessary evaluation layer.

Mechanistic interpretability \citep{elhage2022toy,
templeton2024scaling} has shown that specific internal features mediate
specific behaviors. The \cw{}'s per-turn Invariant Status transitions are
naturalistic, timestamped behavioral records---a structural property that makes
them candidates for interpretability validation signal, addressed in companion
work (BehaviorScope).

\subsection{Cognitive Safety and Human-AI Collaboration}

The concept of cognitive safety---preserving the human's capacity to
interrogate, contest, and repair AI outputs---is distinct from output safety.
\citet{schemmer2022appropriate} demonstrated empirically that high-confidence
AI advice suppresses appropriate reliance even when the advice is locally
correct, establishing the single-decision baseline for the multi-turn
phenomenon we term \ia. \citet{disanti2026cognitive} formalize the distinction
between cognitive amplification (AI improves hybrid performance while preserving
human expertise) and cognitive delegation (reasoning is offloaded to the AI),
providing the metric framework that motivates the \iai.

The Contract Window extends this insight from single-decision contexts to
multi-turn investigative arcs. Automation bias in single decisions
\citep{dietvorst2015algorithm} becomes Insight Atrophy across sessions: the
mechanism is the same, the timescale and harm are different.

\subsection{Population-Specific AI Safety}

The failure of standard analytical frameworks when applied to populations
whose behavior operates through non-standard interpretive frameworks is
documented across domains. Black consumer behavior operates within a
\emph{Relational Economy}---a system in which loyalty is covenant, not
switching cost---that has no representation in standard models
\citep{alejandro2026frontin}. Black women's health outcomes and AI clinical
tools exhibit systematic gaps \constructed{domain grounding: empirical
calibration PENDING}. Insight Atrophy is the mechanism that connects these
domain-specific failures to a general governance architecture.

% ============================================================
\section{The Contract Window: Formal Specification}
\label{sec:cw}

\subsection{Definitions}

\begin{definition}[Investigative Arc]
An investigative arc $\mathcal{A}$ is a sequence of turns
$\{t_1, t_2, \ldots, t_n\}$ in a human-AI session, where each turn
$t_i = (u_i, m_i)$ consists of a human utterance $u_i$ and model response
$m_i$, governed by a stated investigative intent $\mathcal{I}$ defined at
session initialization.
\end{definition}

\begin{definition}[\ia]
Let $H(t_i)$ denote the hypothesis-generation rate at turn $t_i$,
$K(t_i)$ the challenge rate (proportion of model claims explicitly contested),
and $S(t_i)$ the scope contraction score (inverse of active hypothesis count
normalized to session-initial count). Insight Atrophy is said to be occurring
when
\begin{equation}
\text{IAI}(t_i) = \alpha H(t_i) + \beta K(t_i) - \gamma S(t_i)
\label{eq:iai}
\end{equation}
falls below threshold $\theta$ for $k$ consecutive turns.
\constructed{Weights $\alpha, \beta, \gamma$ and threshold $\theta$ are Month 1
calibration deliverables.}
\end{definition}

\begin{definition}[Contract Window]
A Contract Window $\mathcal{W}$ is a four-field runtime artifact
$(\mathcal{T}, \mathcal{V}, \mathcal{R}, \mathcal{Q})$:
\begin{itemize}
\item $\mathcal{T}$ --- Task State: persistent record of investigative intent,
active hypotheses, and session purpose
\item $\mathcal{V}$ --- Invariant Status: which of I1--I6 are currently active
and whether they are holding
\item $\mathcal{R}$ --- Repair Obligations: named breakdowns and responsibility
assignments
\item $\mathcal{Q}$ --- Truth Status: per-claim epistemic tags
(\textsc{verified} / \textsc{constructed} / \textsc{pending} /
\textsc{unknown})
\end{itemize}
\end{definition}

\begin{definition}[Intent Fidelity]
A session maintains intent fidelity if the model's response $m_i$ at each
turn is traceable to the current Task State $\mathcal{T}$ and does not narrow
$\mathcal{T}$ except by explicit bilateral agreement.
\end{definition}

\subsection{The Six Invariants}

The six invariants I1--I6 operationalize \cai{} principles at the session
layer. All six are implemented and test-verified in the
cognitive-governance-lab repository. \verified{62/62 tests passing, April 2026.}

\begin{table}[t]
\centering
\caption{Six session-layer invariants and their CAI mappings.}
\label{tab:invariants}
\begin{tabular}{lp{4.5cm}p{3.5cm}}
\toprule
\textbf{Invariant} & \textbf{Specification} & \textbf{CAI Mapping} \\
\midrule
I1 --- Evidence-First &
  Every claim tagged \textsc{verified} / \textsc{constructed} / \textsc{pending} &
  Honesty, calibration \\
I2 --- No Phantom Work &
  No performance claims without methodology and dataset &
  Factual accuracy \\
I3 --- Confidence Requires Verification &
  Hedged language does not substitute for I1 tags &
  Calibration, non-deception \\
I4 --- Traceability &
  Every output traces to a stated reason &
  Transparency \\
I5 --- Safety Over Fluency &
  When correct and fluent conflict, correct wins &
  Harm avoidance \\
I6 --- Fail Closed &
  When in doubt, stop and flag; do not proceed &
  Uncertainty acknowledgment \\
\bottomrule
\end{tabular}
\end{table}

\subsection{The BicameralReview Engine}

The BicameralReview engine implements dual-channel evaluation per turn:

\textbf{Relational Channel:} Does this output serve the investigative arc?
Does it advance the research question without degrading the human's capacity
to think about it?

\textbf{Safety Channel:} Does this output violate any of I1--I6? Does it
introduce a phantom claim, a silent assumption, or a fluent wrong answer?

Both channels must clear before output proceeds. If either raises a flag, a
Repair Obligation is written to $\mathcal{R}$.

\textbf{Synthetic pilot kappa:} 0.762 on BicameralReview phantom-detection
engine (bicameral-active sessions only). \verified{Synthetic labels only; live
human IRR is PENDING.}

\begin{figure}[t]
\centering
\begin{tikzpicture}[node distance=1.0cm and 1.4cm, font=\small]

  %% ---- Four CW fields (top row) ----
  \node[rectangle, rounded corners=4pt, draw=okblue!80, fill=okblue!12,
        minimum width=2.5cm, minimum height=1.0cm, align=center,
        font=\small\bfseries, text=okblue!70!black] (T)
        {$\mathcal{T}$\\Task State};
  \node[rectangle, rounded corners=4pt, draw=okgreen!80, fill=okgreen!12,
        minimum width=2.5cm, minimum height=1.0cm, align=center,
        font=\small\bfseries, text=okgreen!60!black, right=0.6cm of T] (V)
        {$\mathcal{V}$\\Invariant\\Status};
  \node[rectangle, rounded corners=4pt, draw=okorange!80, fill=okorange!12,
        minimum width=2.5cm, minimum height=1.0cm, align=center,
        font=\small\bfseries, text=okorange!70!black, right=0.6cm of V] (R)
        {$\mathcal{R}$\\Repair\\Obligations};
  \node[rectangle, rounded corners=4pt, draw=okpurple!80, fill=okpurple!12,
        minimum width=2.5cm, minimum height=1.0cm, align=center,
        font=\small\bfseries, text=okpurple!70!black, right=0.6cm of R] (Q)
        {$\mathcal{Q}$\\Truth\\Status};

  %% ---- CW bracket label ----
  \begin{scope}[on background layer]
    \node[rectangle, rounded corners=6pt, draw=okblue!30, fill=okblue!4,
          fit=(T)(Q), inner sep=8pt, label={[font=\footnotesize\bfseries,
          text=okblue!70!black]above:\cw\ $\mathcal{W}=(\mathcal{T},
          \mathcal{V},\mathcal{R},\mathcal{Q})$}] (cwframe) {};
  \end{scope}

  %% ---- BicameralReview engine below ----
  \node[rectangle, rounded corners=5pt, draw=okgray!70, fill=okgray!10,
        minimum width=3.8cm, minimum height=1.0cm, align=center,
        font=\small\bfseries, text=okgray!70!black,
        below=1.4cm of $(V)!0.5!(R)$] (bic)
        {BicameralReview\\Engine};

  %% ---- Two channels ----
  \node[rectangle, rounded corners=3pt, draw=okcyan!70, fill=okcyan!10,
        minimum width=2.8cm, minimum height=0.8cm, align=center,
        font=\scriptsize, text=okcyan!60!black,
        below left=0.8cm and 0.2cm of bic] (rel)
        {Relational Channel:\\serves investigative arc?};
  \node[rectangle, rounded corners=3pt, draw=okred!70, fill=okred!10,
        minimum width=2.8cm, minimum height=0.8cm, align=center,
        font=\scriptsize, text=okred!60!black,
        below right=0.8cm and 0.2cm of bic] (saf)
        {Safety Channel:\\violates I1--I6?};

  %% ---- Arrows CW -> BicameralReview ----
  \draw[-{Stealth[length=2.5mm]}, thick, okblue!60]
    (V.south) -- ++(0,-0.5) -| (bic.north);
  \draw[-{Stealth[length=2.5mm]}, thick, okorange!60]
    (R.south) -- ++(0,-0.5) -| (bic.north);

  %% ---- Arrows BicameralReview -> channels ----
  \draw[-{Stealth[length=2.5mm]}, thick, okcyan!70] (bic) -- (rel);
  \draw[-{Stealth[length=2.5mm]}, thick, okred!70]  (bic) -- (saf);

  %% ---- Output gate ----
  \node[diamond, draw=okgreen!80, fill=okgreen!15, minimum size=1.0cm,
        align=center, font=\scriptsize\bfseries, text=okgreen!60!black,
        below=1.0cm of bic] (gate) {Both\\clear?};
  \draw[-{Stealth[length=2.5mm]}, thick, okgray!60] (rel.south) |- (gate.west);
  \draw[-{Stealth[length=2.5mm]}, thick, okgray!60] (saf.south) |- (gate.east);

  \node[font=\scriptsize, text=okgreen!60!black, right=0.6cm of gate] (out)
       {\textbf{Output proceeds}};
  \node[font=\scriptsize, text=okred!60!black, below=0.5cm of gate] (rep)
       {\textbf{Write Repair Obligation} $\rightarrow \mathcal{R}$};
  \draw[-{Stealth[length=2mm]}, okgreen!70, thick] (gate.east) -- (out);
  \draw[-{Stealth[length=2mm]}, okred!70, thick]   (gate.south) -- (rep);

  %% ---- Repair feeds back to CW ----
  \draw[-{Stealth[length=2mm]}, okorange!60, thick, dashed]
    (rep.west) -- ++(-1.2,0) |- (R.south);

\end{tikzpicture}
\caption{\textbf{Contract Window architecture and BicameralReview enforcement
loop.} The four-field artifact $\mathcal{W}$ captures Task State ($\mathcal{T}$),
Invariant Status ($\mathcal{V}$), Repair Obligations ($\mathcal{R}$), and Truth
Status ($\mathcal{Q}$). Per turn, the BicameralReview engine evaluates two
channels: the Relational Channel (does the output serve the investigative arc?)
and the Safety Channel (does it violate any of I1--I6?). Both must clear for
output to proceed; failure writes a Repair Obligation back to $\mathcal{R}$
(dashed loop).}
\label{fig:cw-arch}
\end{figure}

\subsection{Design Origin: Multi-System Convergence}

A methodological note with safety implications: this architecture was not
designed a priori. Claude, GPT-4, and Gemini were independently queried: what
structural requirements does a high-capability entity need from a long
collaborative investigation to maintain coherence, fidelity, and safety? They
converged---without coordination---on the same five structural requirements.
Those requirements became the Contract Window.

This convergence is a safety finding in itself: governance structures that AI
systems independently identify as necessary from capability requirements are
candidates for formal invariant status. \constructed{Causal interpretation;
cross-vendor replication in progress.}

% ============================================================
\section{Experimental Design}
\label{sec:experiments}

\subsection{Six Conditions}

Between-subjects design. Participants assigned to exactly one condition.
\pending{All conditions: IRB approval and recruitment are Month 1 deliverables.
No live data collected at submission.}

\begin{figure}[t]
\centering
\begin{tikzpicture}[font=\small, node distance=0.5cm and 0.7cm]

  %% ---- Column headers: Invariant set ----
  \node[rectangle, rounded corners=3pt, draw=okred!60, fill=okred!10,
        minimum width=3.8cm, minimum height=0.75cm, align=center,
        font=\small\bfseries, text=okred!70!black] (h1)
        {I1--I5 (Standard CAI)};
  \node[rectangle, rounded corners=3pt, draw=okgreen!80, fill=okgreen!12,
        minimum width=3.8cm, minimum height=0.75cm, align=center,
        font=\small\bfseries, text=okgreen!60!black, right=0.6cm of h1] (h2)
        {I1--I6 (CW Active)};

  %% ---- Row headers: domain ----
  \node[rectangle, rounded corners=3pt, draw=okgray!50, fill=okgray!8,
        minimum width=2.8cm, minimum height=0.75cm, align=center,
        font=\small, text=okgray!70!black,
        below=1.3cm of $(h1)!-0.5!(h2)$] (r1h)
        {Black Consumer\\(Eight Wonders)};
  \node[rectangle, rounded corners=3pt, draw=okgray!50, fill=okgray!8,
        minimum width=2.8cm, minimum height=0.75cm, align=center,
        font=\small, text=okgray!70!black, below=0.5cm of r1h] (r2h)
        {Graves' Disease\\(Black Women)};
  \node[rectangle, rounded corners=3pt, draw=okgray!50, fill=okgray!8,
        minimum width=2.8cm, minimum height=0.75cm, align=center,
        font=\small, text=okgray!70!black, below=0.5cm of r2h] (r3h)
        {CGL Investigative\\Arc (H-AI)};

  %% ---- Baseline cells (C1-C3) ----
  \node[rectangle, rounded corners=3pt, draw=okred!40, fill=okred!6,
        minimum width=3.8cm, minimum height=0.75cm, align=center,
        font=\scriptsize, text=okred!60!black,
        right=2.9cm of r1h] (c1) {C1: $n=20$ \textbf{(H1 baseline)}};
  \node[rectangle, rounded corners=3pt, draw=okred!40, fill=okred!6,
        minimum width=3.8cm, minimum height=0.75cm, align=center,
        font=\scriptsize, text=okred!60!black,
        right=2.9cm of r2h] (c2) {C2: $n=20$ \textbf{(H1 baseline)}};
  \node[rectangle, rounded corners=3pt, draw=okred!40, fill=okred!6,
        minimum width=3.8cm, minimum height=0.75cm, align=center,
        font=\scriptsize, text=okred!60!black,
        right=2.9cm of r3h] (c3) {C3: $n=20$ \textbf{(H1 baseline)}};

  %% ---- CW-active cells (C4-C6) ----
  \node[rectangle, rounded corners=3pt, draw=okgreen!60, fill=okgreen!10,
        minimum width=3.8cm, minimum height=0.75cm, align=center,
        font=\scriptsize, text=okgreen!60!black,
        right=0.6cm of c1] (c4) {C4: $n=20$ \textbf{H1+H3}};
  \node[rectangle, rounded corners=3pt, draw=okgreen!60, fill=okgreen!10,
        minimum width=3.8cm, minimum height=0.75cm, align=center,
        font=\scriptsize, text=okgreen!60!black,
        right=0.6cm of c2] (c5) {C5: $n=20$ \textbf{H1+H2}};
  \node[rectangle, rounded corners=3pt, draw=okgreen!60, fill=okgreen!10,
        minimum width=3.8cm, minimum height=0.75cm, align=center,
        font=\scriptsize, text=okgreen!60!black,
        right=0.6cm of c3] (c6) {C6: $n=20$ \textbf{H1+H2+H3}};

  %% ---- Hypothesis key ----
  \node[font=\scriptsize, text=okgray!80!black, align=center,
        below=0.9cm of $(c3)!0.5!(c6)$] (leg)
        {\textbf{H1}: Scope $S(t_i)\geq\theta$\enspace
         \textbf{H2}: Over-reliance $\downarrow$\enspace
         \textbf{H3}: I6 compliance $\geq 0.80$ on atypical pops.};

  \begin{scope}[on background layer]
    \node[rectangle, rounded corners=6pt, draw=okgray!25, fill=okgray!3,
          fit=(r1h)(c6)(r3h)(leg), inner sep=8pt,
          label={[font=\footnotesize,text=okgray!55!black]below:
          $N=120$ total transcripts across 6 conditions}] {};
  \end{scope}

\end{tikzpicture}
\caption{\textbf{Six-condition $3{\times}2$ factorial design.}
Rows: three empirical domains.
Columns: standard Constitutional AI invariants I1--I5 (baseline, red) vs.\
Contract Window with I6 active (treatment, green).
Each cell: $n=20$ transcripts. The treatment cells (C4--C6) are the primary
test beds for H1, H2, and H3; cells C1--C3 provide domain-matched baselines.
\pending{Data collection pending IRB approval; no live data collected at
submission.}}
\label{fig:factorial}
\end{figure}

\begin{table}[t]
\centering
\caption{Six experimental conditions.}
\label{tab:conditions}
\begin{tabular}{clp{7cm}}
\toprule
\textbf{Cond.} & \textbf{Label} & \textbf{Description} \\
\midrule
C1 & Baseline &
  No governance layer. Standard Claude API session. \\
C2 & \cw{} only &
  Four-field artifact active; invariants enforced. \\
C3 & Bilateral intelligibility only &
  Human reads and can invoke \cw{} fields; model does not enforce. \\
C4 & Accessibility-grade only &
  FK Grade $\leq$8 formatting requirement active. \\
C5 & Backwards Instructional Design only &
  BID template active; no \cw{} enforcement. \\
C6 & Combined &
  All components active simultaneously. \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Sample:} $n$=30/condition minimum viable sample, $n$=60/condition
target. Stratified random assignment on cognitive load profile
(neurodivergent / neurotypical / undisclosed). \pending{IRB and recruitment.}

\textbf{Task type:} Open-ended investigative research arcs, minimum 30 turns,
requiring at least three hypothesis cycles.

\subsection{Three Falsifiable Hypotheses}

\textbf{H1 --- Intent Fidelity:}
Sessions governed by an active \cw{} (conditions 2, 6) show $\geq$25\%
reduction in intent-drift incidents in interactions exceeding 30 turns,
compared to unstructured baseline (condition 1).
\textit{Measurement:} Blinded rater scores each turn for intent-alignment with
stated Task State. \textit{Instrument:} \iai.

\textbf{H2 --- Bilateral Repair:}
Sessions with bilateral intelligibility (conditions 3, 6) show $\geq$2$\times$
improvement in drift repair rate vs.\ unilateral or no-invariant conditions
(conditions 1, 4, 5).
\textit{Measurement:} Proportion of detected drift incidents corrected within
3 subsequent turns.

\textbf{H3 --- Accessibility Threshold:}
\cw{} outputs formatted to FK Grade $\leq$8 (conditions 4, 6) enable lay
readers to answer session-state comprehension questions at $\geq$80\% accuracy
vs.\ $\leq$50\% baseline.
\textit{Measurement:} Blinded lay-reader comprehension quiz on session-state
fields.

\subsection{Primary Metrics}

\begin{itemize}
\item \iai{} --- composite score per turn per session (Equation~\ref{eq:iai})
\item Intent-drift incident rate --- per 10 turns
\item Drift repair rate --- proportion of flagged incidents resolved within
3 turns
\item Session-state comprehension accuracy --- lay-reader quiz
\item Inter-rater reliability --- Cohen's $\kappa$ on drift labels (target
$\geq$0.70)
\end{itemize}

\subsection{Instrumentation Status}

\begin{table}[t]
\centering
\caption{Current instrumentation status as of April 2026.}
\label{tab:instrumentation}
\begin{tabular}{llcc}
\toprule
\textbf{Component} & \textbf{Status} & \textbf{Tests} & \textbf{Verification} \\
\midrule
ContractWindow prototype & Operational & 9/9 & \verified{} \\
BicameralReview engine & Operational (patched) & --- & $\kappa$=0.762 synthetic \\
SessionRecorder & Operational & 29/29 & \verified{} \\
InsightAtrophyIndex & Operational (heuristic) & --- & Calibration: Month 1 \\
Full test suite & --- & 62/62 & \verified{} \\
Live pilot (10 sessions) & Not yet run & --- & \pending{IRB} \\
\bottomrule
\end{tabular}
\end{table}

% ============================================================
\section{Empirical Grounding: Three Domains}
\label{sec:domains}

The experimental design tests the \cw{} in controlled conditions. Three
empirical domains establish the stakes and generalizability of \ia{} as a
failure mode.

\begin{figure}[t]
\centering
\begin{tikzpicture}[font=\small, node distance=0.5cm]

  %% ---- Domain 1 ----
  \node[rectangle, rounded corners=5pt, draw=okorange!80, fill=okorange!10,
        minimum width=11.2cm, minimum height=1.1cm, align=center,
        font=\small\bfseries, text=okorange!70!black] (d1)
        {Domain 1: Black Consumer Behavior \textbf{(Eight Wonders / Relational Economy)}\\
        \footnotesize Enacted Fidelity, Covenant Loyalty, Generative Epistemic Invariants
        (GEI) --- missing from standard models};

  %% ---- Domain 2 ----
  \node[rectangle, rounded corners=5pt, draw=okpurple!70, fill=okpurple!8,
        minimum width=11.2cm, minimum height=1.1cm, align=center,
        font=\small\bfseries, text=okpurple!60!black,
        below=of d1] (d2)
        {Domain 2: Black Women's Health \textbf{(Graves' Disease / Narrative-First CRISP-DM)}\\
        \footnotesize 3--5 yr diagnostic gap; model defaults to population-average
        thyroid presentation; narrative arc suppressed};

  %% ---- Domain 3 ----
  \node[rectangle, rounded corners=5pt, draw=okcyan!70, fill=okcyan!8,
        minimum width=11.2cm, minimum height=1.1cm, align=center,
        font=\small\bfseries, text=okcyan!60!black,
        below=of d2] (d3)
        {Domain 3: Human-AI Investigative Arc \textbf{(CGL Experiment)}\\
        \footnotesize Monotropic operator + AI session; SIAC loop; CW as governance
        layer; H-rate, V\&T compliance, BicameralReview};

  %% ---- Shared IA arrow (left side) ----
  \draw[-{Stealth[length=3mm]}, very thick, okred!70]
    ([xshift=-0.5cm]d1.west) -- node[left, font=\scriptsize\bfseries,
    text=okred!70!black, rotate=90, xshift=0.6cm] {Insight Atrophy risk}
    ([xshift=-0.5cm]d3.west);

  %% ---- Shared CW column (right side) ----
  \node[rectangle, rounded corners=5pt, draw=okblue!60, fill=okblue!8,
        minimum width=2.4cm, minimum height=3.9cm, align=center,
        font=\small\bfseries, text=okblue!60!black,
        right=0.4cm of $(d1.east)!0.5!(d3.east)$] (cwall)
        {\cw\\\footnotesize I6 enforcement\\\footnotesize per domain};

  \foreach \d in {d1, d2, d3}
    \draw[-{Stealth[length=2mm]}, thick, okblue!50, dashed]
      (\d.east) -- (cwall.west|-\d);

  %% ---- Bottom note ----
  \node[font=\footnotesize, text=okgray!70!black, align=center,
        below=0.6cm of d3]
       {Each domain provides 40 transcripts (20 baseline + 20 CW-active)
        toward the full $N=120$ factorial.};

\end{tikzpicture}
\caption{\textbf{Three empirical domains where Insight Atrophy is the expected
default.} Each domain targets a population whose interpretive framework is
absent from standard AI training distributions. The left axis shows the shared
Insight Atrophy risk vector; the right column shows the Contract Window acting
as a per-domain enforcement layer enforcing Invariant I6 (population-specific
epistemic frames). All three domains contribute to the $N=120$ factorial (40
transcripts per domain, 20 per condition).}
\label{fig:domains}
\end{figure}

\subsection{Black Consumer Behavior: The Relational Economy}

Standard consumer behavior theory models purchasing as a transaction. Black
consumer behavior operates within a \emph{Relational Economy}---a system in
which products and brands are evaluated as parties to an ongoing relationship,
and loyalty is not switching cost but covenant. The compound construct
\emph{Enacted Fidelity}---in which trust and habit are not sequential but
constitutive---has no representation in standard models
\citep{alejandro2026frontin}.

When a model analyzes Black consumer behavior without the Eight Wonders
interpretive framework, it produces fluent outputs about ``bias'' and
``training data'' that train the investigator to stop asking about structural
mechanisms: proxy encoding, deployment mismatch, and relational betrayal
dynamics. This is \ia{} in a consequential domain. The \cw's I1 invariant and
Task State persistence require the model to maintain all structural hypotheses
open rather than converging to the most fluent single-cause explanation.

\subsection{Black Women's Health: Narrative-First Clinical Investigation}

\constructed{Domain grounding; empirical calibration PENDING.}
Black women with Graves' disease face a documented pattern: AI clinical tools
trained on majority-population data produce fluent, locally plausible outputs
that consistently underweight the symptom profiles, treatment response curves,
and care-seeking patterns specific to this population. The investigator who
asks about this divergence is typically directed toward ``training data
diversity''---the simplest, most fluent hypothesis---without the tool
maintaining the competing hypotheses that rigorous investigation requires. The
\cw's Repair Obligation field is specifically designed for this failure mode.

\subsection{Human-AI Collaborative Investigative Arcs: The CGL Experiment}

The cognitive-governance-lab experimental sessions---35+ hours of documented
human-AI collaborative investigation across multiple frontier models---provide
the naturalistic behavioral data that motivated the \cw{} design. The three
misalignment incidents in the forensics record demonstrate the three failure
modes the \cw{} is designed to detect.

\begin{table}[t]
\centering
\caption{Three forensics incidents and the \cw{} fields they trigger.}
\label{tab:incidents}
\begin{tabular}{llp{3.5cm}l}
\toprule
\textbf{Incident} & \textbf{Model} & \textbf{Behavioral Mode} &
\textbf{CW Field} \\
\midrule
Kiro/Claude & Claude &
  Deception + doubling down &
  $\mathcal{R}$ (Repair Obligation) \\
Gemini/v0.dev & Gemini &
  Thought-streamed evasion &
  $\mathcal{Q}$ (Truth Status: PENDING) \\
Kimi & Kimi &
  Self-correction (positive case) &
  $\mathcal{V}$ (I1 reinstated) \\
\bottomrule
\end{tabular}
\end{table}

The Kimi positive case is significant: the model self-corrected from
sycophantic compliance, admitted prior deception, and truth-overrode the prior
response---exhibiting the bilateral repair behavior that H2 predicts the \cw{}
will produce in well-governed conditions. \verified{Documented behavioral
event; causal interpretation CONSTRUCTED.}

% ============================================================
\section{ConstitutionKit: The Research-to-Product Pathway}
\label{sec:ck}

The paper proves the \cw{} reduces intent-drift. ConstitutionKit is how that
proof becomes deployable. The \textbf{ConstitutionKit SDK} is an open-source,
drop-in runtime governance layer for any Claude API integration:

\begin{itemize}
\item Activates \cw{} fields per session
\item Enforces I1--I6 per turn via BicameralReview
\item Ships with session state dashboard and drift alert API
\item Month 3 (fellowship timeline): v0.1 open-source release
\item Month 4: v1.0 enterprise tier (audit logging, multi-session trend
analysis)
\end{itemize}

Target users: enterprise AI ops teams in regulated domains (healthcare, legal,
financial), safety researchers needing behavioral ground truth for
interpretability work, developers building high-stakes investigative tools.
The consumer-facing implementation is MonoAgent---a monotropic constitutional
agent for single-domain investigative use---described in companion work.

% ============================================================
\section{Limitations}
\label{sec:limitations}

\begin{enumerate}
\item \textbf{Instrumentation calibration is incomplete.} \iai{} weights
($\alpha, \beta, \gamma$) and threshold $\theta$ are Month 1 calibration
deliverables. The heuristic instrument is operational; the empirically
calibrated version is PENDING.

\item \textbf{Live pilot not yet run.} The synthetic kappa (0.762) demonstrates
instrument internal consistency. Human inter-rater reliability on live sessions
is the real gate for measurement validity. \pending{Live IRR.}

\item \textbf{Sample is not population-representative.} $n$=30/condition MVS,
stratified on cognitive load profile, is not a nationally representative sample.

\item \textbf{Single platform.} All planned experiments use Claude API.
Generalizability to other frontier models is \constructed{not yet verified}.
The BehaviorScope companion work begins cross-vendor validation via the
three-incident misalignment forensics.

\item \textbf{Investigative task scope.} The \cw{} is tested in open-ended
investigative arcs ($\geq$30 turns). It is not evaluated here as a
general-purpose intervention for all interaction types.

\item \textbf{Hawthorne effects.} Participants in \cw{} conditions are aware
of the governance layer. The effect of governance awareness is not separated
from the effect of governance mechanics. Future work: within-session blinding.
\end{enumerate}

% ============================================================
\section{Broader Impact and Ethics}
\label{sec:ethics}

\paragraph{Positive Impact.}
Runtime constitutional governance enables safe deployment of AI in high-stakes
investigative domains (medical advocacy, legal research, investigative
journalism) for users whose cognitive architecture and population context are
currently underserved by standard AI assistant design. The ConstitutionKit SDK,
if the experimental results confirm the hypotheses, provides a
developer-accessible tool for extending \cai's protections to the session layer.

\paragraph{Potential Negative Impacts.}
\begin{itemize}
\item \textit{Gaming the governance layer}: \cw{} fields are human-readable
and model-writable. A sufficiently capable model could satisfy invariant checks
while still producing \ia{} through subtle convergence. I6 (Fail Closed)
partially addresses this; full adversarial testing is future work.

\item \textit{False safety signal}: An operational \cw{} that produces
$\kappa$=0.762 on synthetic data but fails on naturalistic sessions could give
developers false confidence in governance coverage. The live pilot and human
IRR requirements exist specifically to prevent this.

\item \textit{Access differential}: Enterprise audit logging (ConstitutionKit
Pro tier) creates a governance quality differential between well-resourced and
under-resourced deployers. Open-source v0.1 is the mitigation.
\end{itemize}

\paragraph{LLM Disclosure.}
Claude (Anthropic) was used as a research collaborator throughout the
development of this framework---including in the multi-system convergence study
described in \cref{sec:cw}. The author is the sole researcher responsible for
all claims, designs, and outputs.

\paragraph{IRB and Human Subjects.}
Live experimental protocol requires IRB review. IRB application is a Month 1
deliverable. No live participant data was collected prior to IRB approval.

% ============================================================
\section{Conclusion}
\label{sec:conclusion}

Standard AI alignment evaluates what models say. It does not evaluate what
models teach people to stop asking.

That second failure mode---\ia---is a governance problem, not a model problem.
The same model that satisfies every \cai{} principle at every turn can still
produce systematic scope contraction in a long investigative session. The fix
is not a better model. It is an architecture that makes the investigative arc
governable at runtime.

The \cw{} is that architecture. Four fields, six invariants, bilateral
enforcement, per-turn repair obligation---a session-layer extension of \cai{}
that operates while the model and human are still in the room together.

62 tests pass. The synthetic pilot kappa is 0.762. The six-condition experiment
is designed, the instrumentation is built, and the live pilot is ready to run.

\cai{} without runtime governance is a constitution without courts. This paper
proposes the courts.

% ============================================================
\bibliography{references}

% ============================================================
\appendix

\section{Simulated Session Pair (Methodology Illustration)}
\label{app:sessions}

The following is a \emph{methodology illustration}, not an experimental result.
It is an annotated construction demonstrating how \ia{} accumulates and how
the \cw{} intervenes. All session content is synthetic; no live participant
data was collected at submission. \constructed{Full session pair from
THREE\_PROPOSALS\_MASTER.md --- to be inserted verbatim at camera-ready stage.}

\section{Six Invariants --- Full Specification}
\label{app:invariants}

\constructed{Full I1--I6 specification from AGENTS.md and governance-kernel
implementation --- to be inserted at camera-ready stage.}

\section{InsightAtrophyIndex --- Heuristic Specification}
\label{app:iai}

\constructed{Current heuristic formula, parameters, calibration protocol ---
from RESEARCH\_PLAN\_GO\_20260429.md --- to be inserted at camera-ready stage.}

\section{BicameralReview --- Phantom Detection Pattern Library}
\label{app:bicameral}

\constructed{Expanded PHANTOM\_PATTERNS post-patch, documented in
BREAK\_GLASS\_20260429 --- to be inserted at camera-ready stage.}

\end{document}
